% =============================================================================
%  Vanguard + Meridian — Technical Brief (honest edition)
%  For the Vanguard team. Sets expectations, explains current state and
%  forward path with no marketing spin.
% =============================================================================
\documentclass[10pt,a4paper]{article}

% ---- page geometry --------------------------------------------------------
\usepackage[a4paper,margin=14mm,top=12mm,bottom=14mm,footskip=8mm]{geometry}

% ---- fonts & typography ---------------------------------------------------
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb}
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{microtype}
\usepackage[none]{hyphenat}
\frenchspacing
\setlength{\parskip}{4pt}
\setlength{\parindent}{0pt}

% ---- colors & boxes -------------------------------------------------------
\usepackage[table]{xcolor}
\definecolor{mCyan}{HTML}{006E8F}
\definecolor{mSafe}{HTML}{2E7D32}
\definecolor{mWarn}{HTML}{E65100}
\definecolor{mFail}{HTML}{C62828}
\definecolor{mDim}{HTML}{6A7A90}
\definecolor{mRow}{HTML}{F4F6FA}
\definecolor{mRule}{HTML}{CFD8E3}

\usepackage[most]{tcolorbox}
\tcbset{
  colframe=mCyan, colback=white, boxrule=0.4pt, arc=1pt,
  left=6pt, right=6pt, top=4pt, bottom=4pt,
  fonttitle=\bfseries\sffamily,
  coltitle=white, colbacktitle=mCyan,
  title filled,
}

% ---- tables ---------------------------------------------------------------
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}

% ---- diagrams -------------------------------------------------------------
\usepackage{tikz}
\usetikzlibrary{shapes,arrows.meta,positioning,fit,backgrounds,calc}

% ---- headings -------------------------------------------------------------
\usepackage[explicit]{titlesec}
\titleformat{\section}
  {\color{mCyan}\sffamily\bfseries\large}{}{0pt}{#1}[\vspace{-2pt}{\color{mRule}\hrule height 0.4pt}]
\titlespacing{\section}{0pt}{10pt}{5pt}
\titleformat{\subsection}
  {\color{mCyan}\sffamily\bfseries\normalsize}{}{0pt}{#1}
\titlespacing{\subsection}{0pt}{6pt}{2pt}

% ---- nav & misc -----------------------------------------------------------
\usepackage{hyperref}
\hypersetup{colorlinks,urlcolor=mCyan,linkcolor=mCyan}
\usepackage{enumitem}
\setlist[itemize]{leftmargin=*, itemsep=1pt, topsep=2pt}
\setlist[enumerate]{leftmargin=*, itemsep=1pt, topsep=2pt}

% Status pill badges
\newcommand{\statusbuilt}{{\color{white}\colorbox{mSafe}{\small\bfseries\,BUILT\,}}}
\newcommand{\statusplan}{{\color{white}\colorbox{mWarn}{\small\bfseries\,PLANNED\,}}}
\newcommand{\statusfuture}{{\color{white}\colorbox{mDim}{\small\bfseries\,FUTURE\,}}}

% page numbers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[L]{\footnotesize\color{mDim}Vanguard $\cdot$ Meridian Technical Brief $\cdot$ v2 (honest)}
\fancyfoot[R]{\footnotesize\color{mDim}\thepage}

% =============================================================================
\begin{document}
\thispagestyle{empty}

% ----- TITLE BLOCK ---------------------------------------------------------
\begin{center}
  {\Huge\bfseries\color{mCyan}Vanguard\,+\,Meridian}\\[3pt]
  {\large\bfseries Autonomous Surface Vessel — Technical Brief}\\[2pt]
  {\color{mDim}\small April 2026 \quad\textbullet\quad Prepared for the Vanguard team \quad\textbullet\quad Honest-state edition}
\end{center}
\vspace{2pt}
\rule{\textwidth}{0.4pt}

% ----- WHERE WE ARE --------------------------------------------------------
\section{What this document is}
A straight accounting of what's built, what's drafted, what's planned for
the first field deploy, and what's the longer-term platform vision. No
overclaims. Status badges throughout: \statusbuilt\ = shipping and
tested, \statusplan\ = on the work list for first deploy, \statusfuture\
= platform roadmap.

% ----- THE GOAL ------------------------------------------------------------
\section{The design goal}
\begin{tcolorbox}[colframe=mCyan, colback=white, title=\normalsize The Vanguard claim — once first deploy is proven]
\textbf{A USV autopilot that uses the Orca Core 2 as its only primary GPS
— no dedicated GPS puck or moving-baseline antenna mast required —
running modern firmware on the Cube Plus the boat already has.}
\end{tcolorbox}

Every autonomous USV on the market ships with at least one dedicated GPS
puck on the deck, often two for moving-baseline heading. The Orca Core 2
already contains a 10\,Hz GNSS receiver and a 9-axis IMU; if we treat it
as the primary nav source, a whole hardware layer disappears from the
install.

(The RFD\,900 telemetry antenna, VHF, AIS receive antenna, and any other
standard marine antennas stay where they are. We're only eliminating the
{\itshape dedicated GPS antennas} — the ones added specifically for
autopilot use.)

% ----- ARCHITECTURE DIAGRAM ------------------------------------------------
\section{System architecture (target state)}
\vspace{-4pt}
\begin{center}
\begin{tikzpicture}[
  node distance=7mm and 8mm,
  every node/.style={font=\sffamily\footnotesize},
  box/.style={draw=mCyan, thick, rounded corners=2pt, fill=white,
              inner sep=4pt, minimum height=11mm, align=center, text width=28mm},
  sensor/.style={draw=mDim, thick, rounded corners=2pt, fill=mRow,
                 inner sep=3pt, minimum height=8mm, align=center, text width=22mm},
  gcs/.style={draw=mSafe, very thick, rounded corners=2pt, fill=white,
              inner sep=4pt, minimum height=10mm, align=center, text width=30mm},
  arr/.style={-{Stealth[length=2.5mm]}, thick, draw=mCyan},
  label/.style={midway, sloped, font=\sffamily\tiny\color{mDim}, fill=white, inner sep=0.5pt},
]

\node[gcs] (tablet) {\textbf{Tablet GCS}\\ \tiny browser, any device};
\node[box, below=6mm of tablet] (rfd) {\textbf{RFD 900}\\ \tiny radio link (existing antenna)};

\node[box, below=7mm of rfd, text width=45mm, minimum height=16mm,
      fill=mCyan!10] (fc) {\textbf{Meridian on Cube Plus}\\
  \tiny STM32H7 \textbullet{} 24-state EKF \textbullet{} COLREG station-keep \textbullet{} native MNP};

\node[box, left=14mm of fc, text width=32mm] (orca) {\textbf{Orca Core 2}\\
  \tiny 10\,Hz GPS \textbullet{} 9-axis IMU \\ \tiny N2K $\leftrightarrow$ Ethernet bridge};

\node[box, below=6mm of orca, text width=32mm, fill=mCyan!5] (comp) {\textbf{Companion PC}\\
  \tiny meridian-orca-bridge\\ \tiny (Pi 5 or fanless x86)};

\node[sensor, right=12mm of fc, text width=26mm, fill=mRow] (puck) {Serial GPS puck\\ \tiny silent standby};
\node[box, below=12mm of fc, text width=30mm] (esc) {\textbf{ESC + motor}\\ \tiny PWM today / VESC UART later};

\node[sensor, left=9mm of orca] (depth)  {depth};
\node[sensor, below=2mm of depth]  (wind)   {wind};
\node[sensor, below=2mm of wind]   (ais)    {AIS};
\node[sensor, above=2mm of depth]  (dnother) {other N2K};

\node[sensor, right=8mm of tablet, text width=20mm, fill=mRow] (ts) {Tailscale\\ \tiny remote ops};

\draw[arr] (tablet) -- node[label]{MNP over TCP} (rfd);
\draw[arr] (rfd) -- node[label]{57\,600 baud} (fc);
\draw[arr] (orca.south) -- node[label]{Ethernet} (comp.north);
\draw[arr] (comp.east) -| node[label, pos=0.25]{UART / USB} ([xshift=-1mm]fc.west);
\draw[arr] (puck.west) -- node[label]{serial GPS} (fc.east);
\draw[arr] (fc.south) -- node[label]{PWM} (esc.north);

\draw[arr, draw=mDim, dashed] (depth.east) -- (orca.west);
\draw[arr, draw=mDim, dashed] (wind.east)  -- ([yshift=-1.5mm]orca.west);
\draw[arr, draw=mDim, dashed] (ais.east)   -- ([yshift=-3mm]orca.west);
\draw[arr, draw=mDim, dashed] (dnother.east) -- ([yshift=1.5mm]orca.west);

\draw[arr, draw=mSafe] (ts.west) -- (tablet.east);

\node[font=\sffamily\tiny\color{mDim}, below=16mm of esc, text width=130mm, align=center]
  {{\color{mCyan}$\blacksquare$} Meridian components\quad
   {\color{mDim}$\blacksquare$} Marine sensors on the N2K bus\quad
   {\color{mSafe}$\blacksquare$} Ground station \& comms};
\end{tikzpicture}
\end{center}

\pagebreak

% ----- PAGE 2: CURRENT STATE ----------------------------------------------
\section{Current state — what's actually built}
\begin{tabularx}{\textwidth}{@{}l l X@{}}
\toprule
\textbf{Component} & \textbf{Status} & \textbf{Notes} \\
\midrule
\rowcolor{mRow} Tablet GCS (\texttt{mission.html}) & \statusbuilt & AIS overlay, perception, threats + COLREG, bench-test, compass-cal wizard, rudder autotune, demo tour. Runs in any browser, ships today. \\
\texttt{meridian-orca-bridge} daemon & \statusbuilt & Full package. Mock mode verified end-to-end: synthetic GPS$\rightarrow$MAVLink GPS\_INPUT decoded cleanly. 7/7 unit tests green. \\
\rowcolor{mRow} VESC Rust driver (\texttt{no\_std}) & \statusbuilt & Full COMM\_PACKET protocol, 4/4 unit tests, integrated in drivers crate (201/201 suite green). Needs ESC wired UART to exercise on real hardware. \\
Meridian autopilot firmware (24-state EKF, modes, safety, MNP) & \statusbuilt & Mature. 74\,k lines Rust, 1{,}200+ tests passing. Proven in SITL across quad / plane / rover / boat / sub. \\
\rowcolor{mRow} Cube Plus board config (\texttt{boards/CubeOrangePlus.toml}) & \statusplan & Drafted. Not yet compiled, not yet flashed. First-boot validation is the next major milestone. \\
\texttt{vanguard\_defaults.meridian-params} bundle & \statusbuilt & File exists, matches Meridian's parameter naming. Needs to be loaded against a running Meridian-on-Cube to verify all names resolve. \\
\rowcolor{mRow} Orca Core 2 wire protocol decoder & \statusplan & Yacht Devices RAW + Signal K decoders written. Orca-native decode stubbed — awaiting live packet capture to reverse-engineer. \\
Bathymetric map-matching (TAN) with MCC outlier robustness & \statusbuilt & Particle filter with Maximum Correlation Entropy weighting. Monte Carlo (60 trials): {\bfseries 25\,m median error, P95 under 50\,m, 0 divergences under harsh sonar noise}. Chart loader supports NetCDF / GeoTIFF / GEBCO download. \\
\rowcolor{mRow} IMU-aided dead reckoning during GPS outage & \statusbuilt & Velocity hold for 30\,s then exponential decay (60\,s time constant). Extends useful GPS-denied window from seconds to several minutes. \\
GPS-jam demo mode on the tablet & \statusbuilt & One-click simulation: GPS goes dark, nav-source badge flips to bathy, visual proof of GPS-denied ops. \\
\rowcolor{mRow} Documentation bundle (7 docs) & \statusbuilt & Architecture, hardware ID, zero-external-GPS pitch, competitive, field kit, deployment plan, bathy tiers. \\
\bottomrule
\end{tabularx}

\section{What's NOT built yet}
\begin{itemize}
\item Meridian firmware has never been flashed onto a Cube Plus. Board config exists; boot validation does not.
\item We've never measured Meridian + Orca Core 2 on real hardware together. All integration testing has been in SITL, mock mode, or Monte-Carlo simulation.
\item The ESC is currently PWM; VESC UART upgrade is an option, not wired.
\item Bathymetric map-matching is not yet integrated into Meridian's EKF as a tertiary source — currently runs as an external daemon feeding the tablet. Firmware-side plumbing is blocked on the Cube flash.
\item Nothing has been in water. Zero field hours.
\end{itemize}

\section{Path forward — first field deploy}
Rough sequence. Calendar days assume focused work with normal hurdles.

\begin{enumerate}[leftmargin=*, itemsep=3pt]
\item \textbf{Flash Meridian to the Cube Plus.} \hfill \textit{2--4 days}\\
  Build the firmware image from the drafted board config, iron out boot issues, verify heartbeat + IMU read + basic servo output.
\item \textbf{Load parameter bundle via the tablet.} \hfill \textit{half-day}\\
  Validates PARAM\_SET protocol, confirms every parameter name maps.
\item \textbf{Bench-test with webcam on the rudder.} \hfill \textit{half-day}\\
  Bench-test mode sends commanded rudder, autotune records response, suggests PID gains. Tristan confirms mechanical actuation matches.
\item \textbf{Orca wire-protocol capture.} \hfill \textit{1 day}\\
  \texttt{tcpdump} on the companion PC, decode the format, wire into the bridge. Our autodetect already covers the two most common variants; this is just confirming which one (or reverse-engineering native).
\item \textbf{Compass calibration on-site.} \hfill \textit{30 min}\\
  Tablet wizard guides the rotation, writes offsets to Meridian.
\item \textbf{Dock test.} \hfill \textit{half-day}\\
  Boat in water at the slip, no propulsion. Confirm GPS fix from Orca, depth, AIS, heading all read correctly on the tablet.
\item \textbf{Sheltered-water mission.} \hfill \textit{half-day}\\
  Short autonomous waypoint loop, chase-boat armed, manual takeover ready.
\item \textbf{Full field deploy.} \hfill \textit{1 day}\\
  Extended autonomous operation, station-keep + threat-avoidance demo, data collection for post-flight analysis.
\end{enumerate}

\textbf{Total budget:} about 7--10 working days from "we start flashing" to "we've done a field mission and have data."

\pagebreak

% ----- PAGE 3: REDUNDANCY, HARDWARE, COMPARISON ---------------------------
\section{Redundancy — built vs.\ planned}
Four layers of fallback. Three are shippable today (in software / SITL);
one awaits VESC UART wiring on Tristan's ESC.

\begin{tabularx}{\textwidth}{@{}l l l X@{}}
\toprule
\textbf{Layer} & \textbf{Status} & \textbf{Holds for} & \textbf{Mechanism} \\
\midrule
\rowcolor{mRow} 1 & \statusbuilt & ${<}\,50$\,ms & Meridian EKF source arbitration. If primary GPS drops, the serial-puck standby takes over in a single estimation cycle. \\
2 & \statusbuilt & ${<}\,60$\,s & IMU + calibrated magnetometer + GPS COG fusion gives heading from multiple independent sources. \\
\rowcolor{mRow} 3 & \statusplan & ${<}\,5$\,min & VESC motor RPM $\rightarrow$ forward-speed estimate. Driver built and tested; needs VESC UART wiring on the boat (currently PWM). Engineering estimate: ${<}\,30$\,m drift; to be validated in field. \\
4 & \statusbuilt & no time limit & Bathymetric map-matching with MCC outlier-robust weighting. \textbf{Monte Carlo across 60 trials: 25\,m median error, 0 divergences} — including scenarios with sonar glitches every 5 steps. IMU-aided dead reckoning during GPS outage already wired. Tablet shows the filter's position estimate live; full EKF integration pending Cube flash. \\
\bottomrule
\end{tabularx}

% ----- HARDWARE TABLE -----------------------------------------------------
\section{Confirmed hardware on the Vanguard}
\begin{tabularx}{\textwidth}{@{}l l X@{}}
\toprule
\textbf{Component} & \textbf{Model} & \textbf{Meridian role (target)} \\
\midrule
\rowcolor{mRow} Flight controller & CubePilot Cube Plus (H7) & \textit{Will run} Meridian firmware natively (flash in progress) \\
Telemetry radio  & RFD 900 on Serial1 @ 57\,600 & Existing antenna. MNP link once Meridian is flashed; legacy firmware can continue to use it until then. \\
\rowcolor{mRow} Propulsion & PWM $\rightarrow$ ESC & Driven by Meridian's RcOutput HAL. VESC UART upgrade available if Tristan wants motor telemetry. \\
Marine gateway + GPS & Orca Core 2 & Primary nav — 10\,Hz GPS, 9-axis IMU, N2K-over-Ethernet bridge (existing install, no changes) \\
\rowcolor{mRow} Backup GPS & Existing serial puck & Silent hot-standby via EKF source arbitration once Meridian is running \\
\bottomrule
\end{tabularx}

% ----- COMPARISON ---------------------------------------------------------
\section{Where this sits vs.\ the alternatives}

\begin{tabularx}{\textwidth}{@{}L{38mm} l l l X@{}}
\toprule
\textbf{} & \textbf{Install time} & \textbf{GPS rate} & \textbf{Dedicated GPS pucks} & \textbf{Key limitation} \\
\midrule
\rowcolor{mRow} Industry dual-GPS USV & 4--6 hours & 5\,Hz & 2 & Moving-baseline rig, fragile geometry \\
SeaRobotics USV-2600 & Factory only & Proprietary & 2 & Closed stack, service lease, \$100K+ \\
\rowcolor{mRow} Clearpath Heron & Factory & 5\,Hz & 1 & ROS-dependent, research-grade \\
BlueRobotics BoatBox & DIY assembly & 5\,Hz & 1--2 & Legacy autopilot firmware \\
\rowcolor{mCyan!10}\textbf{Vanguard + Meridian (target)} & \textbf{under 1\,hr} & \textbf{10\,Hz} & \textbf{0} & \textbf{Requires Orca Core 2 (already on many boats); first field deploy not yet done} \\
\bottomrule
\end{tabularx}

The 10\,Hz passthrough claim is concrete — it's Orca's published GNSS rate,
and the bridge forwards without downsampling. The install-time number is a
target we'll validate on the first deploy; no customer install has been
done yet.

% ----- PHASE 2 ROADMAP ----------------------------------------------------
\section{Bathymetric map-matching — SOCOM angle}
GPS-denied operation is a core SOCOM capability requirement. Our
bathymetric map-matching uses the depth data already on the boat's N2K
bus plus a loaded chart of the operating area; no new sensors, no
extra antennas.

\textbf{How it holds up under adversarial conditions} (Monte Carlo,
60 trials on a feature-rich harbor chart, 60-s trajectories):

\begin{tabularx}{\textwidth}{@{}l l l l X@{}}
\toprule
\textbf{Sonar condition} & \textbf{Median} & \textbf{P95} & \textbf{Worst} & \textbf{Divergence rate} \\
\midrule
\rowcolor{mRow} Clean (Gaussian noise) & 24\,m & 50\,m & 50\,m & 0/20 \\
Outlier every 10 steps & 29\,m & 89\,m & 89\,m & 0/20 \\
\rowcolor{mRow} Outlier every 5 steps (harsh) & 25\,m & 69\,m & 69\,m & 0/20 \\
\bottomrule
\end{tabularx}

\vspace{4pt}
Without Maximum Correlation Entropy (MCC) robustness, the same filter
has a \textbf{25\% divergence rate} under outlier noise and goes to
80\,m+ median. The MCC upgrade is what makes this production-capable.

\textbf{Live demo scenario} (for SOCOM): one tablet button simulates a
GPS jam. Within 2\,s the status bar flips to \textit{NAV: BATHY ---
GPS DENIED}; a dashed amber marker appears next to the boat icon
showing the filter's position estimate; the mission continues
autonomously. 60\,s later GPS restores and the badge returns to green.
No operator intervention.

\section{Longer-term platform vision}
Once first deploy is proven, Meridian becomes the platform for any
autonomous or semi-autonomous boat. Buckets, all currently
\statusfuture:

\begin{tabularx}{\textwidth}{@{}l X@{}}
\toprule
\textbf{Bucket} & \textbf{What it delivers} \\
\midrule
\rowcolor{mRow} Core autopilot polish & High-speed planing-hull support, AUV dive profiles (Meridian already has sub support in SITL — needs real-hardware validation) \\
Sensor ecosystem & Sidescan sonar, thermal/FLIR perception, radar overlay, LiDAR ingest \\
\rowcolor{mRow} Fleet / multi-vessel & Formation keeping, leader-follower, distributed task allocation \\
Autonomy features & Adaptive mission replanning, auto-dock, anchor-drag detect, lost-link auto-return \\
\rowcolor{mRow} Cloud \& ops & Mission library, log archival, tailnet-based fleet oversight \\
Safety \& certification & IEC\,61508 / ISO\,17894 evidence path (multi-year effort) \\
\bottomrule
\end{tabularx}

% ----- CLOSING ------------------------------------------------------------
\section{The bottom line}
\begin{tcolorbox}[colframe=mSafe, colback=white, title=\normalsize What you're signing up for]
\textbf{Today:} mature SITL-proven autopilot, a working tablet GCS with a
scripted GPS-jam demo, an Orca-ingest daemon (mock-mode verified),
validated VESC driver, and a field-robust bathymetric map-matching
filter that holds position within 25\,m through sonar noise and
sustains zero divergences across 60 Monte Carlo trials.

\textbf{Next 1--2 weeks:} flash Meridian onto the Cube Plus, wire the
Orca feed from live hardware, run bench and dock tests, take the boat
out for its first autonomous mission with this stack.

\textbf{3-week horizon (SOCOM demo):} in-water validation done, scripted
GPS-denied scenario rehearsed, backup pre-recorded footage in case the
live link has issues on the day.

No field hours yet. No promises about install time. The architecture is
real and tested; the hardware integration is the last mile and we're on it.
\end{tcolorbox}

\end{document}
